An Examination of Sport Climbing Competition Format and Scoring System

### 2021-07-30 manuscript/Rmd/draft.Rmd
Some edits.

In 2016, the International Olympic Committee announced the addition of five new sports to the 2020 Summer Olympics in Tokyo, Japan, which would then reschedule for 2021 due to the impact of the COVID-19 global pandemic. The five new events added to Tokyo 2020's competition program were baseball/softball, karate, skateboard, sports climbing, and surfing [@ioc2016]. One of the new sports, sport climbing, got our attention, specifically because of its unique scoring system and the fact that only one set of medals is awarded for each gender.

Sport climbing at the 2020 Tokyo Olympics consists of three disciplines: speed climbing, bouldering, and lead climbing. Speed climbing takes place on a standardized course and competitors try to reach the top of the course as fast as possible. For Tokyo 2020, speed climbing is being contested in a head-to-head format with ranks determined by how far a competitor advances in the bracket. In bouldering, contestants have a fixed amount of time to complete as many courses as they can. Winners are determined based on who completes the most courses and ties are broken based on who had the fewest attempts. Ties are further broken by the competitor achieved the most "zone holds", which are holds approximately halfway through each course. Finally, in lead climbing, an athlete gets one point for each hold that they reach, so whoever reaches the highest point on the wall is the winner. Each lead climber only gets one attempt and when they fall their attempt is over. These three different climbing disciplines demand different sets of skills and, often, athletes specialize in a single event. However, since only one set of Olympic medals is awarded to sport climbing, rather than choosing only one of these disciplines to include in the Olympics, all three events were chosen to be included as a sort of climbing triathlon.

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world. In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint". She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

>>>>

In 2016, the International Olympic Committee (IOC) announced the addition of five new sports to the 2020 Summer Olympics in Tokyo, Japan, which would then reschedule for 2021 due to the impact of the COVID-19 global pandemic. The five new events added to Tokyo 2020's competition program were baseball/softball, karate, skateboard, surfing, sports climbing [@ioc2016]. One of the new sports, sport climbing is particularly interesting because of its scoring system which uses the product of ranks across three disciplines to determine the medalists, with the lowest rank product declared the winner.

#While there are three main disciplines in sport climbing, the IOC only allotted one set of medals per gender.  So rather than choosing only a single discipline, they chose to combine all three of the events together to create a combined sport climbing event.

The three disciplines that comprise sport climbing at the 2020 Tokyo Olympics are: speed climbing, bouldering, and lead climbing. Speed climbing takes place on a standardized course and competitors try to reach the top of the course as fast as possible. For Tokyo 2020, speed climbing is being contested in a head-to-head format with ranks determined by how far a competitor advances in the bracket. In bouldering, contestants have a fixed amount of time to complete as many courses as they can. Winners are determined based on who completes the most courses and ties are broken based on who had the fewest attempts. Ties are further broken by the competitor achieved the most "zone holds", which are holds approximately halfway through each course. Finally, in lead climbing, an athlete gets one point for each hold that they reach, so whoever reaches the highest point on the wall is the winner. Each lead climber only gets one attempt and when they fall their attempt is over.

These three different climbing disciplines demand different sets of skills and, often, athletes specialize in a single event. However, since only one set of Olympic medals is awarded to sport climbing, rather than choosing only one of these disciplines to include in the Olympics, all three events were chosen to be included as a sort of climbing triathlon.

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world (CITATION NEEDED). In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint". She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.
